---
title: "Introduction to GenoSim: Forward-Time Genotype Simulation"
author: "GenoSim Authors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to GenoSim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE
)
library(GenoSim)
```

# Overview

**GenoSim** addresses a critical bottleneck in clinical and population
genetics research: the shortage of human NGS genotype data due to ethical,
budgetary, and regulatory constraints. It provides a complete forward-time
simulation pipeline that produces biologically realistic diploid SNP data
across up to 10 generations.

Two simulation modes are available:

| Mode | Function | When to use |
|---|---|---|
| **Population** | `simulate_population()` | No prior data; generate from allele frequencies |
| **Pedigree** | `simulate_from_pedigree()` | Real family VCFs + pedigree structure available |

---

# Installation

```{r install, eval=FALSE}
# From a local zip file:
install.packages("GenoSim_1.1.3.tar.gz", repos = NULL, type = "source")

# Or via remotes from GitHub (once published):
# remotes::install_github("example/GenoSim")
```

---

# Population Mode

## Basic Simulation

The simplest use case: generate 100 founder individuals with 1,000 SNPs and
run 5 generations of neutral evolution.

```{r pop-basic}
sim <- simulate_population(
  n_founders    = 100,
  n_snps        = 1000,
  n_generations = 5,
  chromosomes   = 1:5,
  seed          = 42,
  verbose       = FALSE
)

# Each list element is a dosage matrix [n_indiv x n_SNPs]
names(sim$genotypes)
dim(sim$genotypes[["gen0"]])
```

## Inspecting Summary Statistics

```{r pop-stats}
sim$summary_stats[, c("generation", "n_individuals",
                       "obs_heterozygosity", "mean_maf", "inbreeding_fis")]
```

The `inbreeding_fis` column tracks the Wright $F_{IS}$ statistic:

$$F_{IS} = 1 - \frac{H_o}{H_e}$$

Under neutral panmictic evolution with no inbreeding, $F_{IS} \approx 0$.

## Consanguineous / High-Inbreeding Scenario

Set `inbreeding_F = 0.125` to model a first-cousin mating system:

```{r pop-inbred}
sim_inbred <- simulate_population(
  n_founders    = 80,
  n_snps        = 1000,
  n_generations = 5,
  inbreeding_F  = 0.125,
  n_eff         = 60,
  chromosomes   = 1:5,
  seed          = 2024,
  verbose       = FALSE
)

sim_inbred$summary_stats[, c("generation","obs_heterozygosity",
                              "exp_heterozygosity","inbreeding_fis")]
```

## Selection + Drift

```{r pop-sel}
sim_sel <- simulate_population(
  n_founders    = 200,
  n_snps        = 2000,
  n_generations = 8,
  selection_s   = 0.05,    # positive selection on alt allele
  n_eff         = 150,
  chromosomes   = 1:10,
  seed          = 999,
  verbose       = FALSE
)
```

## Visualisation

```{r pop-viz, fig.height=5}
plot_dashboard(sim_inbred, pca_result = run_pca(
  do.call(rbind, sim_inbred$genotypes), n_pc = 5
))
```

```{r af-traj}
plot_af_trajectory(sim_inbred, n_snps_show = 30)
```

---

# Analysis Functions

## Hardy-Weinberg Equilibrium Testing

```{r hwe}
hwe <- hwe_test(sim_inbred$genotypes[[1]], alpha = 0.05)
table(hwe$sig_label)
```

## Linkage Disequilibrium

```{r ld}
ld <- compute_ld(
  sim$genotypes[[1]], sim$snp_map,
  max_snps = 200, max_dist_bp = 5e6
)
cat(sprintf("Pairs: %d | Mean r²: %.4f\n", nrow(ld), mean(ld$r2, na.rm=TRUE)))
```

```{r ld-plot}
plot_ld_decay(ld)
```

## Runs of Homozygosity

```{r roh}
roh <- detect_roh(
  sim_inbred$genotypes[[6]],
  sim_inbred$snp_map,
  min_snps      = 10,
  min_length_bp = 500000
)
nrow(roh$roh_segments)
```

```{r roh-plot}
plot_roh_per_individual(roh)
```

## PCA

```{r pca}
all_genos <- do.call(rbind, sim$genotypes)
pca <- run_pca(all_genos, n_pc = 10)
cat(sprintf("PC1: %.1f%%  PC2: %.1f%%  PC3: %.1f%%\n",
            pca$variance_pct[1], pca$variance_pct[2], pca$variance_pct[3]))
```

```{r pca-plot}
gen_labels <- sub("_ind.*", "", rownames(pca$scores))
plot_pca(pca, color_by = gen_labels)
```

## FST and Diversity

```{r fst}
fst <- compute_fst(sim$genotypes)
print(fst)
```

```{r diversity}
div <- diversity_metrics(sim)
print(div)
```

---

# Pedigree Mode

## Loading Example Data

GenoSim ships a complete example dataset: 25-member family **FAM_KHAN** across
4 generations with 4 affected individuals and consanguineous matings.

```{r ped-load}
ped <- read_pedigree(example_ped_path(), verbose = FALSE)
summarise_pedigree(ped)
```

```{r vcf-load}
vcf <- read_vcf_cohort(example_vcf_dir(), verbose = FALSE)
cat(sprintf("Loaded: %d individuals × %d SNPs\n",
            nrow(vcf$geno_matrix), ncol(vcf$geno_matrix)))
```

## Running Pedigree-Constrained Simulation

```{r ped-sim}
ped_sim <- simulate_from_pedigree(
  vcf_cohort        = vcf,
  pedigree          = ped,
  extra_generations = 3,
  mut_rate          = 1e-4,
  seed              = 42,
  verbose           = FALSE
)

ped_sim$summary_stats[, c("generation","n_individuals",
                           "obs_heterozygosity","mean_maf","source")]
```

## Pedigree Visualisations

```{r ped-tree, fig.width=10, fig.height=7}
affected <- ped$individual_id[ped$phenotype == 2]
plot_pedigree_tree(ped, highlight_ids = affected,
                   title = "FAM_KHAN — Consanguineous Pedigree")
```

```{r kinship, fig.width=7, fig.height=6}
plot_kinship_heatmap(ped)
```

```{r af-compare}
plot_af_comparison(vcf, ped_sim)
```

---

# Exporting Results

## VCF Export

```{r export-vcf, eval=FALSE}
export_vcf(sim, generation = c(0, 5), out_dir = "output/vcf/")
```

## PLINK Export

```{r export-plink, eval=FALSE}
export_plink(sim, generation = 5, out_prefix = "my_sim", out_dir = "output/plink/")
```

## CSV Export

```{r export-csv, eval=FALSE}
export_csv(sim, out_dir = "output/csv/")
# Produces:
#   genosim_genotypes_all_generations.csv
#   genosim_summary_stats.csv
#   genosim_allele_freqs.csv
#   genosim_snp_map.csv
```

---

# Session Information

```{r session}
sessionInfo()
```
